---
title: "us_tuition_costs"
author: "Justin Shotwell"
date: "4/22/2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}

library(tidyverse)
library(janitor)
library(lubridate)
library(ggthemes)
library(scales)

theme_set(theme_tufte())

```

This is the average cost for "one year of full-time study at public four-year institutions" by state.

```{r data, include=FALSE, message=FALSE}

us_tuition <- readxl::read_excel("us_avg_tuition.xlsx") %>%
  clean_names() %>%
  pivot_longer(-state, names_to = "year", values_to = "cost") %>%
  mutate(year = str_replace(year, "x", ""),
         year = str_replace(year, "_.*", ""),
         year = as.integer(year) - 2004)

```

The first thing I think to ask is how have the states changed relatvie to other states.

```{r warning=FALSE, message=FALSE}

us_tuition %>%
  ggplot(aes(year +  2004, cost, group = state)) +
  geom_line() +
  geom_smooth(method = "lm", group = 1) +
  labs(x = "Year",
       y = "One Year of Tuition")

```
```{r warning=FALSE, message=FALSE}

us_tuition %>%
  filter(state == "Illinois") %>%
  ggplot(aes(year +  2004, cost, group = state)) +
  geom_line() +
  geom_smooth(method = "lm", group = 1) +
  labs(x = "Year",
       y = "One Year of Tuition")

```


```{r warning=FALSE, message=FALSE}

us_tuition %>%
  filter(state == "Georgia") %>%
  ggplot(aes(year +  2004, cost, group = state)) +
  geom_line() +
  geom_smooth(method = "lm", group = 1) +
  labs(x = "Year",
       y = "One Year of Tuition")

```

That just tells us that the states are all rising in cost over time. I suspect that some states are rising at different rates. One way to see this clearer would be to create a simple linear model for each state that compares year and cost. Then sort the results by the highest coefficient estimate.

```{r}

by_state <- us_tuition %>%
  group_by(state) %>%
  nest()

state_model <- function(df) {
  lm(cost ~ year, data = df)
}
  
by_state <- by_state %>% 
  mutate(model = map(data, state_model))

by_state_tidy <- by_state %>% 
  mutate(tidy = map(model, broom::tidy)) %>% 
  unnest(tidy) %>%
  select(state, term, estimate) %>%
  filter(term == "year") %>%
  ungroup() %>%
  mutate(state = as_factor(state))

str(by_state_tidy)

```

```{r}

by_state_tidy %>%
  mutate(state = fct_reorder(state, estimate)) %>%
  arrange(desc(estimate)) %>%
  head(10) %>%
  ggplot(aes(state, estimate)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "College Cost Rate of Change by State",
       subtitle = "Top 10 Rates of Increase",
       x = "",
       y = "Average Estimated Change in Cost per Year")

```

```{r}

by_state_tidy %>%
  mutate(state = fct_reorder(state, estimate)) %>%
  arrange(desc(estimate)) %>%
  tail(10) %>%
  ggplot(aes(state, estimate)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "College Cost Rate of Change by State",
       subtitle = "Bottom 10 Rates of Increase",
       x = "",
       y = "Average Estimated Change in Cost per Year")

```

